\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}

### 2024-04-10 doc/paper/supplements.tex
Updates from Overleaf

\newcommand{\supp}{SI Appendix}

\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}

% author list
\author[1$\ddag$]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[1]{Kathrin Zürcher}
\author[2]{Robin Wood}
\author[1]{Matthias Egger}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}

@@

\affil[$\ddag$]{These authors contributed equally to this work.}

@@

\section{Study setting and data}\label{sec:setting-and-data}

\subsection{Building}

\begin{figure}[!htpb]
    \centering
    \includegraphics{doc/paper/floor_plan_2021.pdf}
    \caption{Floor plan showing the defined areas inside the clinic and the placement of the video sensors and CO$_2$ devices.}
    \label{fig:floor-plan}
\end{figure}

\subsection{Tracking data}

\begin{figure}[!htpb]
    \centering
    \includegraphics{illustrations/tracking-examples.pdf}
    \caption{Sample tracks from our video sensors of this study.}
    \label{fig:tracking-examples}
\end{figure}

\subsection{Preprocessing of tracking data}

Tracking data required preprocessing before it could be fully used for spatiotemporal modeling. The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database. Clinical attendees most likely contributed multiple tracks for several reasons: 
\begin{itemize}
    \item A person moved outside the range of the video sensors and then re-entered. 
    \item An individual may temporarily not have been recognized as a person by the tracking software, \eg because the person bowed.
    \item A person may have been temporarily hidden behind another person.
    \item Moving objects such as opening doors may have been recognized as a person, adding a few noisy tracks into the data.
\end{itemize}
We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person. In the first step, we filtered noise (\ie IDs of less than 10s and a total moving distance of less than 2m) and automatically matched IDs that were less than 10s and 3m apart while moving (\ie IDs anywhere in the clinic) or less than 60s and 0.5m apart while sitting (\ie two IDs ending/starting in a seating area of the clinic). In the second step, we build on the automatically matched data and performed an additional manual linkage. The aim was to link IDs that were further apart but still most likely belonged together. Through manual inspection, we could better assess whether a link resulted in a movement that made sense and preclude better alternatives. A Shiny application and a matching manual/wiki (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) was developed to assist the students that performed the manual linkage.

The Shiny app is shown in \Cref{fig:shiny-app}. In the right panel, a floor plan of the clinic is shown in (x,y)-space. Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue. Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track. The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey. Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link. Possible links are sorted in ascending order by the time difference.

In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information, \ie the name of the loaded file at the top and the path to the file where the updated data is automatically saved at the bottom, the date of the tracking, the initial number of IDs before any linkages and current number of IDs after linkages, and the number of links made. We created labels to denote finished tracks as "Ended" if the ID was continuously tracked throughout the clinic or "Lost" if the ID was somewhere lost in the clinic. The number of labels are shown below the number of IDs. Below the descriptive information are the tools related to performing linkages. First, the currently selected ID (black track) is shown, together with action buttons allowing students to quickly move to the first unfinished ID, the next or previous ID, or the previous or next ID starting at the entrance. Second, students could decide whether they want to match forward or backward in time. Third, the duration and number of links are shown for the current ID, which for example can alert students about IDs spending an unusually long time in the clinic. Fourth, the sliding bars allow students to filter possible links below a certain difference in time, distance, or height. Pre-set filters are added atop to quickly switch between common thresholds. Fifth, there is a drop-down menu to select IDs for which possible alternative links should be shown (light grey tracks). If there is a more plausible alternative, students could select the corresponding ID and check whether it is indeed the better match. Sixth, another drop-down menu below selects the ID which should be linked to the current patient ID (colored tracks) and the action "Link IDs" performs the link and updates the floor plan, table, and all other related information. If the students realized that the link was erroneous, they could unlink the last/first ID. Finally, once no further links could be made, either because the ID left the clinic or was lost, students could finish the track and label it accordingly. Ended tracks are removed from the set of possible links whereas lost tracks remain in the set.

\begin{figure}[!htpb]
    \centering
    \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
    \caption{Screenshot of the Shiny app used for manual linking IDs that probably belong together.}
    \label{fig:shiny-app}
\end{figure}

@@

\subsection{The Wells-Riley model}

We extend the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection. The model estimates the probability of airborne indoor transmission $P$ using the following equation: 
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
\end{align}
where $C$ is the number of disease cases, $S$ is the number of susceptibles, $I$ is the number of infectious individuals in space, $p$ is the breathing rate per person, $q$ is the generation rate of infectious quanta, $t$ is the duration of exposure, and $Q$ is the outdoor air supply rate.

@@

While considering the stochastic behaviour of airborne transmission, the Wells-Riley model makes two simplifying assumptions. First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and evenly in the indoor airspace. In other words, the exposure to infectious quanta is the same regardless of the locations of the infectious and susceptible individual. Second, the Wells-Riley model assumes steady-state conditions, which means that the quanta concentration and the outdoor air supply rate are constant over time. In other words, the risk of infection for each susceptible depends only on the duration of exposure (the time they spent in the indoor space).

Rudnick and Milton\cite{Rudnick2003IndoorAir} modified the Wells-Riley model to relax the steady-state assumption. They use CO$_2$ as a biomarker for exhaled breath to compute the outdoor air supply rate, which is otherwise difficult to measure. Their modified equation is
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{\bar{f}Iqt}{n}\right),
\end{align}
where $f = \frac{CO^{\text{I}}-CO^{\text{O}}}{CO^{\text{A}}}$ is the fraction of indoor air that is exhaled breath, which is computed based on the CO$_2$ level in indoor air $CO^{\text{I}}$, outdoor air $CO^{\text{O}}$, and exhaled breath $CO^{\text{A}}$ (in parts per million [ppm]).

While the extension by Rudnick and Milton resolves the steady-state assumption, it still assumes a well-mixed airspace. Therefore, our aim is to extend the Wells-Riley model further to allow the risk of infection to vary both across time and airspace. We accomplish this by modeling the quanta concentration spatiotemporally. The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points. Formally, we estimate the cumulative risk of infection for susceptible individual $a$ as 
\begin{equation}\label{eq:spattemp-P}
    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
\end{equation}
where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether $a$ was at location $s$ at time $t$, and $p$ is the breathing rate per person (in $m^3/s$). The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta: 
\begin{itemize}
    \item[\ref{sec:quanta-generation}] Quanta generation: Infectious individuals generate quanta at their location in the indoor space. 
    \item[\ref{sec:quanta-diffusion}] Quanta diffusion: Over time, the infectious quanta diffuse and we approach a well-mixed airspace. 
    \item[\ref{sec:quanta-removal}] Quanta removal: Infectious quanta is removed from the air through (a)~outdoor air exchange, (b)~viral inactivation, and (c)~gravitational settling. 
\end{itemize}
We describe each process in more detail in the following. While we apply our model to \emph{Mtb}, we note that it can be applied to airborne respiratory infections in general.

@@

We divide the airspace in a 2-dimensional grid, irrespective of the height of the airspace. The cell denotes the unit airspace with $s = 1, \dots, S$, where $(x_s, y_s)$ are the x and y coordinate of cell $s$. In some cases, we may use the (x,y)-notation in place of the cell ID $s$. The quanta concentration will be the same within each grid cell $s$, thus the grid provides a discrete approximation to the spatially varying quanta concentration. The number of grid cells $S = n_x \cdot n_y$ corresponds to the size of the matrix and determines the quality of the approximation as well as the model run time. More grid cells (smaller grid cells) provide a more granular picture of the spatial quanta concentration, at the expense of higher computation time.

We define the following variables and modeling parameters:
\begin{itemize}
    \item $N_{s,t}$: quanta concentration in airspace unit $s$ at time $t$
    \item $I_{s,t}$: number of infectious individuals in airspace unit $s$ at time $t$
    \item $q$: quanta generation rate (in quanta/s)
    \item $D$: diffusion constant (in m$^2$/s)
    \item $AER$: air change rate (in air changes/s)
    \item $V$: volume of the airspace (in m$^2$)
    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air
    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air
    \item $G$: average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$)
    \item $n_t$: number of individuals in airspace at time $t$
    \item $\lambda$: viral inactivation rate
\end{itemize}

@@

The number of quanta generated by infectious individuals at time $t$ is 
\begin{align}\label{eq:generation}
    I_t \cdot q ~.
\end{align}
The initial spatial spread of the quanta over the grid is informed prior literature. The concentration of virus-laden aerosols is highest closest to the source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}, in line with findings showing that the risk of infection correlates with the distance to the infectious individual\cite{Ko2004RiskAnal,Kenyon1996NEJM}. Although coughing and sneezing can transport aerosols further away from the individual, these activities are much less frequent than breathing. Therefore, prior work suggests that breathing may contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}. Overall, it seems plausible that the majority of infectious aerosols are generated near the infectious individual and then disperse over time. Since we do not observe the directional flow of the exhaled breadth, we assume that the generated quanta is evenly and radially distributed around the individual.

@@

Absent further knowledge on airflow, we assume that aerosols diffuse uniformly in the indoor air, over time approaching a well-mixed airspace. Initially, the quanta concentration will be highest at the location where it is generated and then disperse radially in (x,y)-direction towards locations where the concentration is lower. The diffusion equation is 
\begin{align}\label{eq:diffusion}
    \frac{\delta N_{s,t}}{\delta t} = D \Delta N_{s,t},
\end{align}
where $\Delta$ is the Laplace operator and $D$ is the diffusion constant (in m$^2$/s). We solve this equation with a standard second-order finite difference model using the Euler method with a 5-point stencil.

The diffusion constant $D$ is unknown but can be approximated using an empirical relationship between the eddy diffusion coefficient $K$ and the outdoor air exchange rate $AER$ (in air changes/h)\cite{Cheng2011EnvSciTech}
\begin{align}
    K = (0.52 \cdot AER + 8.61\cdot10^{-5}) \cdot V^{\frac{2}{3}},
\end{align}
where $V$ is the volume of the airspace (in m$^3$). We assume that $D \approx K$.

The $AER$ can be estimated from indoor CO$_2$ levels and room occupancy. Under steady-state conditions (CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
\begin{align}
    AER = \frac{6\cdot10^4 \cdot n \cdot G}{V\cdot(CO^{\text{I}}-CO^{\text{O}})},
\end{align}
where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person (in L$\cdot$min$^{-1}\cdot$person$^{-1}$), $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air (both in parts per million [ppm])\cite{Batterman2017IJERPH}. In a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration. Therefore, we determine $AER$ using a transient mass balance model of the form\cite{Batterman2017IJERPH}
\begin{align}
    CO_{t+1}^{\text{I}} = \frac{6\cdot10^4 \cdot n_t + G}{Q} \cdot \left(1 - \exp(-Q/V \Delta t)\right) + (CO_t^{\text{I}}-CO^{\text{O}}) \cdot \exp(-Q/V \Delta t) + CO^{\text{O}},
\end{align}
where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$. We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) as implemented in the statistical software R\cite{Byrd1995SIAM}, a quasi-Newton method. The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error. We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [350,500]$\cite{Batterman2017IJERPH}. The $AER$ may be computed at different time intervals, thus we add the subscript $t$, \ie $AER_t$.

@@

Infectious quanta may be removed from the air through outdoor air exchange, viral inactivation, or gravitational settling. We will ignore gravitational settling. Vuorinen (2020)\cite{Vuorinen2020SafSci} notes that particles smaller than 20\,$\mu$m rapidly become airborne. This is far above the upper end of the particle size distribution of most respiratory infections, including \emph{Mtb}, which is carried in particles typically smaller than 4.7\,$\mu$m\cite{Fennelly2020Lancet}. The removal rate is thus the same of the outdoor air exchange rate $AER_t$ and the viral inactivation rate $\lambda$, \ie 
\begin{align}\label{eq:removal}
    AER_t + \lambda ~.
\end{align}
Note that $\lambda$ depends on the pathogen and is generally difficult to estimate and likely also depends on environmental factors, particularly relative humidity [Reference].

@@

We illustrate our modeling approach with an example. Consider a room with a volume $V$=150m$^3$ (length [L] = 10m, breadth [B] = 5m, height [H] = 3m). We divide the room into a grid of $S = 40 \cdot 20 = 800$ cubic cells with dimensions 0.25m$^2 \cdot 3$m, \ie a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m. We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells. We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle of the room, \ie cell (20,10), so that the generation of quanta will be confined to cells $(x_s = 20\pm1, y_s = 10\pm1)$, \ie $N_{s,t} = (I \cdot q) / 9 ~~ \forall s \in (x_s = 20\pm1, y_s = 10\pm1)$. We further assume a quanta generation rate of $q = 10$\,quanta/h, a constant outdoor air exchange rate of $AER = 1$\,air changes/h (corresponding to a removal rate of 1 and a diffusion coefficient of 0.007m$^2$/s), and a viral inactivation rate of $\lambda = 0.5$.

\Cref{fig:toy-example} shows the quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am). The quanta is generated at the location of the infectious individual. The quanta concentration remains high around this location until the individual ceases to generate more quanta at 9am. The quanta diffuses relatively quickly and at 10min past 9am the airspace is well-mixed with quanta. Over time, the removal through outdoor air exchange and viral inactivation brings the concentration in the room back towards 0.

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../tests/stm_v2-toy_example.pdf}
    \caption{Quanta concentration at different time points following the generation of the first quanta (8:00:00am) and last quanta (9:00:00am) from an infectious individual that is placed in a fixed location in the middle of the room with dimensions L=10m $\times$ B=5m $\times$ H=3m.}
    \label{fig:toy-example}
\end{figure}

@@

\section{Estimation of infection risk in the primary care clinic}\label{sec:estimation}

\subsection{Building and setup}\label{prep:building}

A schematic view of the clinic is shown in \Cref{fig:clinic}. Clinical attendees were tracked when entering the clinic, in the waiting room, the corridor, and the TB room. If they moved into one of the care rooms along the corridor or into other parts of the clinic they were only tracked once they came back. We divide the clinic into three spaces: the waiting room (L=10.55m $\times$ B=5.50m $\times$ H=3.00m), the corridor (L=12.45m $\times$ B=2.20m $\times$ H=2.50m), and the TB room (L=4.75m $\times$ B=3.50m $\times$ H=3.00m). We separated the waiting room and the corridor because we observed small differences in the time-varying CO$_2$ levels, but we acknowledge that the two areas could also be evaluated jointly since the doors to the corridor were often most of the time. Moreover, the corridor and waiting room are not completely enclosed, but connected to the entrance, exit, and other parts of the clinic. Thus, we simplify our modeling analysis when we limit quanta diffusion within these rooms and also note that the calculated air change rate is affected by the separation of the waiting room and corridor (\ie the volume and room occupancy would otherwise be different).

>>>>

\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risk in a South African primary care clinic using environmental, clinical, and patient movement data}}

% author list
\author[1]{Nicolas Banholzer}
\author[2]{Keren Middelkoop}
\author[2]{Juane Leukes}
\author[3]{Ernest Weingartner}
\author[1]{Remo Schmutz}
\author[1]{Kathrin Zürcher}
\author[1]{Matthias Egger}
\author[2]{Robin Wood}
\author[1*]{Lukas Fenner}

\affil[1]{Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland}
\affil[2]{Desmond Tutu HIV Centre, Department of Medicine, University of Cape Town, Cape Town, South Africa}
\affil[3]{Institute for Sensors and Electronics, University of Applied Sciences and Arts Northwestern Switzerland, Windisch, Switzerland}

@@

@@

\section{Processing of tracking data}\label{sec:setting-and-data}

Tracking data required processing before it could be fully used for modeling. The reason is that there were considerably more tracking IDs than what would be expected from the number of registered patients in the clinical database. Visitors of the clinic most likely contributed multiple tracks for several reasons: 
\begin{itemize}
    \item A person moved outside the range of the video sensors and then re-entered. 
    \item An individual may temporarily not have been recognized as a person by the tracking software because the person bowed downwards.
    \item A person may have been temporarily hidden behind another person.
    \item Moving objects such as opening doors may have been recognized as a person, introducing noise.
\end{itemize}
We proceeded in two steps to link back together tracking IDs that most likely belonged to the same person. In the first step, we filtered noise (\ie IDs of less than 10\,s and a total moving distance of less than 2\,m), and then automatically matched the remaining tracks that were less than 10\,s and 3\,m apart while moving or less than 60\,s and 0.5\,m apart while sitting. In the second step, we build on the automatically matched tracking data and, through manual inspection, linked tracks that were further apart but still most likely of the same person. To facilitate the manual linkage, we developed a Shiny application and a matching manual (\url{https://github.com/nbanho/clinic-transmission/wiki/Matching-Manual}) that assisted the students who were performing this linkage.

The Shiny app is shown in \Cref{fig:shiny-app}. In the center panel, a floor plan of the clinic is shown in (x,y)-space. Different areas of the clinic are delineated by colors, \eg seating areas are marked in purple and exit areas are marked in blue. Tracks are shown as lines with an open circle indicating the start and a crossed circle indicating the end of a track. The currently selected track is shown in black, possible links are shown in colors (here only one possible link shown in red), and alternatives to the selected possible link are shown in light grey.

Below the floor plan is a table showing descriptive information about the possible links in relation to the currently selected track: (1)~the patient ID (Pid) of the selected track and observation ID (Oid) of the possible link, (2)~the height difference between the end of the Pid and end of the Oid track, (3)~ the standing height (\ie maximum height of a track) difference, (4)~the duration of the possible link, (5)~the time difference, and (6)~the distance between the end of the selected track and the start of the selected possible link. Possible links are sorted in ascending order by the time difference.

In the left panel of \Cref{fig:shiny-app}, the Shiny app first shows descriptive information: the loaded file, the date of the tracking data, the initial number of IDs before any manual linkages, the current number of IDs after considering manual linkages, and the current number of made linkages. It then shows the current Pid with shortcuts to move to the next or previous Pid, or the next or previous Pid at the entrance. Linkages can be done either forward or backward in time. The duration of the current Pid (time in the clinic) is shown together with the number of linked raw tracks of which the current Pid consists. The sliding bars allow filtering of possible links according to difference in time, distance, or height, with pre-set filters above to quickly switch between common thresholds. The first dropdown menu selects the Oid for which alternative matches should be shown in the floor plan. The idea of this feature is to prevent false linkages when there are more plausible alternatives. A linkage is made by selecting the Oid in the "Link with IDs" dropdown menu and clicking "Link IDs". After that, the app updates the floor plan, table, and information in the panel. If an erroneous link was made, it can be revoked by clicking "Unlink last ID" if the last link in time should be revoked or by clicking "Unlink first ID" if the first link in time should be revoked. If no further linkages could be made, the track can be finished by entering a label (see below). The manually linked dataset is automatically saved to the file shown in the bottom of the left panel.

We created labels to denote finished tracks: 
\begin{itemize}
    \item \textbf{Clean}: A clean track throughout the clinic with known entrance and exit.
    \item \textbf{Staff}: IDs with moving in- and out of the staff room.
    \item \textbf{Noise}: A short track ($\leq$1\,min) without entrance or exit.
    \item \textbf{Lost enter}: A long track (>1\,min) without entrance.
    \item \textbf{Lost exit}: A long track (>1\,min) without exit.
    \item \textbf{Lost both}: A long track (>1\,min) without entrance and without exit.
\end{itemize}

\begin{figure}[!htpb]
    \centering
    \includegraphics[width=\linewidth]{doc/paper/shiny-tool-screenshot.png}
    \caption{Screenshot of the Shiny app used for manual linking tracking IDs that probably belong together.}
    \label{fig:shiny-app}
\end{figure}

After manual processing of the tracking data, we arrived at 3,319 tracks. We then filtered noise and short tracks <5\,min to obtain 1,563 used for analysis: 209 (13\%) "Clean", 125 (8\%) "Staff", 213 (14\%) "Lost enter", 246 (16\%) "Lost exit", and 770 (49\%) "Lost both". A sample of our tracks used for analysis, excluding staff, are shown in \Cref{fig:tracking-examples}.

\begin{figure}[!htpb]
    \centering
    \includegraphics{results/data/example-patient-tracks.png}
    \caption{Sample tracks from our video sensors of this study.}
    \label{fig:tracking-examples}
\end{figure}

@@

\subsection{The Wells-Riley modeling framework}

We build upon the Wells-Riley model\cite{Riley1978AJE}, which is an established epidemiological model to estimate the risk of infection. The model estimates the risk (probability) of airborne indoor transmission $P$ using the following equation: 
\begin{align}
    P = \frac{D}{S} = 1 - \exp \left(-\frac{Ipqt}{Q}\right),
\end{align}
where $D$ is the number of diseased cases, $S$ is the number of susceptible cases, $I$ is the number of infectious people in the indoor space, $p$ is the breathing rate (m$^3$ h$^{-1}$), $q$ is the quantum (infectious dose) generation rate (quanta h$^{-1}$), $t$ is the exposure time (s), and $Q$ is the ventilation rate (m$^3$ h$^{-1}$).

@@

While considering the stochastic behaviour of airborne transmission, the Wells-Riley model makes two simplifying assumptions. First, it assumes a well-mixed airspace, which means that the generated quanta disperse immediately and evenly in the indoor airspace. In other words, the exposure to infectious quanta is the same regardless of the location of the infectious and susceptible individual. Second, the Wells-Riley model assumes steady-state conditions, which means that the quanta concentration and the outdoor air supply rate are constant over time. In other words, the risk of infection for each susceptible only varies with the duration of exposure (the time they spent in the indoor space).

Rudnick and Milton\cite{Rudnick2003IndoorAir} modified the Wells-Riley model to relax the steady-state assumption. They use CO$_2$ as a biomarker for exhaled breath to compute the outdoor air supply rate, which is otherwise difficult to measure. Their modified equation is
\begin{align}
    P = \frac{C}{S} = 1 - \exp \left(-\frac{\bar{f}Iqt}{n}\right),
\end{align}
where $f = \frac{CO^{\text{I}}-CO^{\text{O}}}{CO^{\text{A}}}$ is the fraction of indoor air that is exhaled breath, which is computed based on the CO$_2$ level in indoor air $CO^{\text{I}}$, outdoor air $CO^{\text{O}}$, and exhaled breath $CO^{\text{A}}$ (in parts per million [ppm], respectively).

While the extension by Rudnick and Milton relaxes the steady-state assumption, it still assumes a well-mixed airspace. We relax this assumption by modeling the concentration of infectious quanta over space time. Then, we compute the cumulative risk of infection for each susceptible individual over space and time as 
\begin{equation}\label{eq:spattemp-P}
    P = \sum_s \sum_t N_{c,t} \cdot \mathbb{I}_{c,t} \cdot p,
\end{equation}
where $N$ is the quanta concentration in cell airspace $c$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether the susceptible individual was in cell $c$ at time $t$, and $p$ is the breathing rate.

Modeling the spatiotemporal quanta concentration involves three processes
\begin{itemize}
    \item[\ref{sec:quanta-generation}] \textbf{Quanta generation}: Quanta is generated at the infectious individual's location. 
    \item[\ref{sec:quanta-diffusion}] \textbf{Quanta diffusion}: Quanta diffuses until we approach a well-mixed airspace. 
    \item[\ref{sec:quanta-removal}] \textbf{Quanta removal}: Quanta is removed from the air or becomes inactive (not infectious). 
\end{itemize}
We describe each process in more detail in the following, but first we introduce some notation.

@@

We divide the area of the airspace in a 2-dimensional grid of cells $c = 1, \dots, C$, where $(x_c, y_c)$ refer to the center of cell $c$. The grid provides a discrete approximation to the continuously varying quanta concentration across the space. The number of grid cells $C = n_x \cdot n_y$ determines the spatial resolution and computation time, \ie more cells increase resolution at the expense of more computation time. We model the quanta concentration at time $t = 1, \dots T$ (in seconds).

We define the following variables and modeling parameters:
\begin{itemize}
    \item $N_{c,t}$: quanta concentration in airspace unit $c$ at time $t$
    \item $I_{c,t}$: number of infectious individuals in airspace unit $c$ at time $t$
    \item $q$: quanta generation rate (quanta s$^{-1}$)
    \item $D$: diffusion constant (m$^2$ s$^{-1}$)
    \item $AER$: air change rate (air changes s$^{-1}$ / quanta s$^{-1}$)
    \item $V$: volume of the airspace (m$^3$)
    \item $CO^{\text{I}}$: CO$_2$ concentration in indoor air (ppm)
    \item $CO^{\text{O}}$: CO$_2$ concentration in outdoor air (ppm)
    \item $G$: average CO$_2$ generation rate per person (L$\cdot$min$^{-1}\cdot$person$^{-1}$)
    \item $n_t$: number of individuals in airspace at time $t$
    \item $\lambda$: bacterial inactivation rate (quanta s$^{-1}$)
    \item $k$: gravitational settling rate (quanta s$^{-1}$)
\end{itemize}

@@

The number of quanta generated by infectious individuals at time $t$ in cell $c$ is 
\begin{align}\label{eq:generation}
    I_{c,t} \cdot q ~.
\end{align}
For now, we assume that the initially generated quanta is confined to the cell where the infectious individuals are located, which is not an unreasonable assumption. The concentration of virus-laden aerosols is typically highest near the infectious source\cite{Vuorinen2020SafSci,Chen2020BuildEnv}. Although coughing and sneezing can expel the aerosols further away from the individual, these activities are much less frequent than breathing, which may thus contribute more infectious particles to the airspace than coughing or sneezing\cite{Dinkele2022AJRCCM}.

@@

Absent measurements of airflow, we assume that aerosols diffuse uniformly in the indoor air, approaching a well-mixed airspace over time. We assume that the quanta disperse radially in (x,y)-direction towards locations where the concentration is lower. The diffusion equation is 
\begin{align}\label{eq:diffusion}
    \frac{\delta N_{c,t}}{\delta t} = D \Delta N_{c,t},
\end{align}
where $\Delta$ is the Laplace operator and $D$ is the diffusion constant. We solve this equation with a standard second-order finite difference model using the Euler method with a 5-point stencil.

The diffusion constant $D$ is unknown. We approximate it using an empirical relationship between the eddy diffusion coefficient $K$ ($K \approx D$) and the outdoor air exchange rate $AER$ for CO$_2$ particles\cite{Cheng2011EnvSciTech,Foat2020BE}
\begin{align}
    K = (0.52 \cdot AER + 8.61\cdot10^{-5}) \cdot V^{\frac{2}{3}},
\end{align}
where $V$ is the volume of the airspace (in m$^3$). While this is only an approximation, it nicely relates the air change rate to quanta diffusion.

The $AER$ is not directly observed, but can be estimated from indoor CO$_2$ levels, room occupancy, and room volume. Under steady-state conditions (\ie CO$_2$ levels reaching a steady-state concentration), $AER$ can be computed as
\begin{align}
    AER = \frac{6\cdot10^4 \cdot n \cdot G}{V\cdot(CO^{\text{I}}-CO^{\text{O}})},
\end{align}
where $n$ is the number of people in the airspace, $G$ is the average CO$_2$ generation rate per person, $CO^{\text{I}}$ is the CO$_2$ concentration in indoor and $CO^{\text{O}}$ is the concentration in outdoor air\cite{Batterman2017IJERPH}. However, in a primary care clinic, room occupancy varies continuously and thus it may be difficult to determine the steady-state CO$_2$ concentration. Therefore, we determine $AER$ using a transient mass balance model\cite{Batterman2017IJERPH}
\begin{align}
    CO_{t+1}^{\text{I}} = \frac{6\cdot10^4 \cdot n_t + G}{Q} \cdot \left(1 - \exp(-Q/V \Delta t)\right) + (CO_t^{\text{I}}-CO^{\text{O}}) \cdot \exp(-Q/V \Delta t) + CO^{\text{O}},
\end{align}
where $Q$ is the outdoor air supply rate (in m$^3$/h) and $AER = Q/V$. We numerically solve the transient mass balance model with the limited memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS), a quasi-Newton method, as implemented in the statistical software R\cite{Byrd1995SIAM}. The optimal solution is the outdoor air supply rate $Q$ that minimizes the root mean square error. We also follow recommendations to fit the outdoor CO$_2$ level with reasonable constraints $CO^{\text{O}} = [300,600]$\cite{Batterman2017IJERPH}. We can compute $AER$ for different time windows, hence $AER_t$.

@@

Infectious quanta may be removed from the air through outdoor air exchange, bacterial inactivation, or gravitational settling. We will ignore gravitational settling because \emph{Mtb} is carried in small particles between 2\,$\mu$m and 5\,$\mu$m\cite{Fennelly2020Lancet}, which become airborne immediately or within seconds\cite{Vuorinen2020SafSci}. The removal rate is thus the sum of the outdoor air exchange rate $AER_t$ and the bacterial inactivation rate $\lambda$ for \emph{Mtb}, \ie 
\begin{align}\label{eq:removal}
    AER_t + \lambda ~.
\end{align}

\subsection{Spatiotemporal quanta concentration}

@@

We illustrate our modeling approach with an example. Consider a room with a volume $V$=150m$^3$ (length = 10m, breadth = 5m, height = 3m), which we rasterize into a grid of $C = 40 \cdot 20 = 800$ cells with an area of 0.25m$^2$. We assume that quanta generation is confined to the cell where the infectious individual is located as well as the first neighbouring cells. We monitor the room from 8am to 12am and consider one infectious individual generating quanta continuously for one hour between 8am and 9am at a fixed location in the middle, \ie cell (20,10), which means that the initially generated quanta is $(x_c = 20\pm1, y_c = 10\pm1)$, \ie $N_{c,t} = \frac{1}{9} \cdot I \cdot q$ in cells $c \in (x_s = 20\pm1, y_s = 10\pm1)$, and zero in all other cells. We further assume a quanta generation rate of $q = 10$\,quanta h$^{-1}$, a constant outdoor air exchange rate of $AER = 1$\,air changes h$^{-1}$ (removal rate of 1\,quanta h$^{-1}$ and corresponding to a diffusion coefficient of 0.007\,m$^2$ s$^{-1}$), and a bacterial inactivation rate of $\lambda = 0.5$\,quanta h$^{-1}$.

\Cref{fig:toy-example} shows the quanta concentration at different time points. The quanta concentration increases over time and is higher near the infectious individual. After the individual has left the space, the quanta diffuses fairly quickly and to approach a well-mixed already at 9:05am. Over time, the removal through outdoor air exchange and bacterial inactivation brings the concentration in the room back towards zero.

\begin{figure}[!htpb]
    \centering
    \includegraphics{tests/stm_v2-toy_example.png}
    \caption{Spatiotemporal quanta concentration between 8am and 10am following the generation of quanta by an infectious individual between 8am and 9am in the middle of the room.}
    \label{fig:toy-example}
\end{figure}

@@

\section{Simulation approach}\label{sec:estimation}

\subsection{Setup}\label{prep:building}

We divide the clinic into three spaces: the waiting room (length = 10.55m $\times$ breadth = 5.50m $\times$ height = 3.00m), the corridor (length = 12.45m $\times$ breadth = 2.20m $\times$ height = 2.50m), and the TB room (length = 4.75m $\times$ breadth = 3.50m $\times$ height = 3.00m). We separate the waiting room and the corridor because separate rectangular rooms simplify modeling and because we observe small differences in the time-varying CO$_2$ levels, suggesting these rooms are somewhat different. However, we note that the doors to the corridor were often most of the time, so that the corridor and waiting room were not completely enclosed. As such, it is possible that aerosolized \emph{Mtb} (quanta) spreads from the waiting room to the corridor, and vice versa, but this is not considered in our modeling approach. We rasterize each room into a grid of cubic cells with a squared area of 0.25m$^2$, corresponding to a diagonal length of $\sqrt{0.25^2 + 0.25^2} = 0.35$\,m. Since the rooms have different height, the quanta concentration in the corridors' cells will (all else equal) be higher than in the waiting and TB room by a factor of 3/2.5 = 1.2.
